Revealing Benefits, Costs, and Tradeoffs of Spatial Management in Marine Systems Arrising from Coupled Economic and Ecological Dynamics

### 2023-04-03 03_marlin_paper.qmd
slight edits to movement notation

We provide a brief overview of the the general CTMC method here, see
@thorson2021a for a detailed description. Under this framework, movement
from each patch to each other patch in the system in a given timestep
*t* for life stage *a* of species *s* is defined by a movement matrix
$\boldsymbol{M}_{t,s,a}$. $\boldsymbol{M}_{t,s,a}$ is calculated as the
matrix exponent of a function of the diffusion $\boldsymbol{D}$ and
taxis $\boldsymbol{\tau}$ matrices scaled by the width of the time step
$\Delta_{t}$ and the length of the edge of each patch $\Delta_d$ that
the model is running under. So for example, if we are running a model on
a monthly timestep, and we define the units of time in years, then
$\Delta_{t} = 1/12$. Similarly, if we are running a model where one
patch equals 100km^2^, then $\Delta_d = 10km$. This parameterization
allows users to set the effective area of the spatial domain through two
avenues; the number of patches, which effectively scales the resolution
of the model, and the area of each patch, which scales the spatial
extent of the simulation.

$$
\boldsymbol{M}_{t,s,a} = e^{\Delta_{t}(\boldsymbol{D}_{t,s,a} + \boldsymbol{\tau}_{t,s,a})} 
$$ {#eq-movement}

The individual components (*m*) of the movement matrix are filled based
on an adjacency matrix, which defines whether two patches are both
adjacent and water (as opposed to land or another physical barrier), a
diffusion rate $D$ defined in units of area of a patch per unit of time,
and a habitat preference function *h* in units of length of a side of a
patch per unit time. For example, if we are defining the time units as
years and the distance units as kilometers, for a tuna $D$ might be
1,000 $\frac{KM^2}{Year}$. We then use parameters $\Delta_{t}$ and
$\Delta_{d}$ parameters to translate the diffusion rate $D$ to match the
time step and patch size used in a simulation. This "scale free"
parameterization means that appropriate value of $D$ can be identified
for a species and then set, regardless of the time step or patch size
used in the simulation model itself. The taxis component of the movement
process is a function of the difference in habitat quality *h*. The
habitat preference function itself can take any form the user wishes.
Exponentiating the difference in the habitat preference function between
patches turns the taxis matrix into a multiplier of the diffusion rate
*D*. As such, when creating habitat layers for simulation, users can
tune the scale of the habitat gradient function to result in realistic
multipliers of the diffusion rate. This parameterization ensure that the
off-diagonal elements of the movement matrix $\boldsymbol{M}_{t,s,a}$
are all non-negative, a requirement of the CTMC method.

$$ 
m_{p1,p2,t,s,a} = \begin{cases}
      = \frac{\Delta_{t}}{\Delta_{d}^2}De^{\frac{\Delta_t(h(p2,t,s,a) - h(p1,t,s,a))}{\Delta_d}} & \text{if p2 and p1 are adjacent}\\
     = -\sum_{p' \neq p1} m_{p1,p2,t,s,a} & \text{if p1 = p2}\\
     = 0 & \text{otherwise.}
\end{cases}
$$ {#eq-diffusion}

@@

>>>>

We provide a brief overview of the the general CTMC method here (see
@thorson2021a for a detailed description). Under this framework, movement of indivuduals
from each patch to each other patch in the system in a given timestep
*t* for life stage *a* of species *s* is defined by a movement matrix
$\boldsymbol{M}_{t,s,a}$. $\boldsymbol{M}_{t,s,a}$ is calculated as the
matrix exponent of a function of the diffusion $\boldsymbol{D}$ and
taxis $\boldsymbol{\tau}$ matrices scaled by the width of the time step
$\Delta_{t}$ and the length of the edge of each patch $\Delta_d$ that
the model is running under. So for example, if we are running a model on
a monthly timestep, and we define the units of time in years, then
$\Delta_{t} = 1/12$. Similarly, if we are running a model where one
patch equals 100km^2^, then $\Delta_d = 10km$. This parameterization
allows users to set the effective area of the spatial domain through two
avenues; the number of patches, which effectively scales the resolution
of the model, and the area of each patch, which scales the spatial
extent of the simulation.

The individual components (*M*) of the movement matrix ($\boldsymbol{M}$) are filled based
on an adjacency matrix, which defines whether two patches are both
adjacent and water (as opposed to land or another physical barrier), a
diffusion rate $D$ defined in units of area of a patch per unit of time,
and a habitat preference function *h* in units of length of a side of a
patch per unit time. For example, if we are defining the time units as
years and the distance units as kilometers, for a tuna $D$ might be
1,000 $\frac{KM^2}{Year}$. We then use parameters $\Delta_{t}$ and
$\Delta_{d}$ parameters to translate the diffusion rate $D$ to match the
time step and patch size used in a simulation. This "scale free"
parameterization means that appropriate value of $D$ can be identified
for a species and then set, regardless of the time step or patch size
used in the simulation model itself. The taxis component of the movement
process is a function of the difference in habitat quality *h*. The
habitat preference function itself can take any form the user wishes.
Exponentiating the difference in the habitat preference function between
patches turns the taxis matrix into a multiplier of the diffusion rate
*D*. As such, when creating habitat layers for simulation, users can
tune the scale of the habitat gradient function to result in realistic
multipliers of the diffusion rate. This parameterization ensure that the
off-diagonal elements of the movement matrix $\boldsymbol{M}_{t,s,a}$
are all non-negative, a requirement of the CTMC method.

$$ 
M_{p1,p2,t,s,a} = \begin{cases}
      = \frac{\Delta_{t}}{\Delta_{d}^2}De^{\frac{\Delta_t(h(p2,t,s,a) - h(p1,t,s,a))}{\Delta_d}} & \text{if p2 and p1 are adjacent}\\
     = -\sum_{p' \neq p1} M_{p1,p2,t,s,a} & \text{if p1 = p2}\\
     = 0 & \text{otherwise.}
\end{cases}
$$ {#eq-diffusion}

@@

The movement of individuals across patches is then calculated by matrix multiplication of the pre-movement vector of the number of individuals ($\boldsymbol{n}$) of species *s* at age *a* in time step *t* across all patches *p* times the matrix exponential of the movement matrix $\boldsymbol{M}$

$$
\boldsymbol{n}_{t+1,s,a} = \boldsymbol{n}_{t,s,a}e^{\boldsymbol{M}_{t,s,a}}
$$ {#eq-movement}
